Towards topological type theory for decrypting transfinite methods in classical mathematics

### 2025-08-25 slides-ai2025.tex
Final version

  \textbf{Theorem.} There is a \hil{unique} best~$L_1$-approximation of~$f$ in~$P_n$.
  \pause

@@

\begin{frame}{Backed by logical metatheorems?}
  \hil{Metatheorems} have been and are being developed which \ldots
  \begin{enumerate}
    \item guarantee the \hil{extractability} of suitable numerical information in principle
    (bounds on norms, convergence rates, moduli of uniqueness, rates of
    asymptotic regularity, \ldots),
    \item describe an \hil{algorithm} for carrying out the extraction and
    \item support \hil{modular treatments} of auxiliary lemmas,
  \end{enumerate}
  \bad{\textbf{provided}} the input proof is supplied in a
  certain \hil{formal system}.
  \pause

  NB: The \hil{quality} of the extracted data depends on the
  \hil{sophistication} of the \hil{logical~principles} used in the proof
  (Heine--Borel, Bolzano--Weierstraß, \ldots).
  \pause

  \bad{\textbf{In practice (2025):}} Algorithms \bad{\textbf{not}} used, only followed
  as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
  \hil{Let us explore tool support!}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-0em}\includegraphics[width=\paperwidth]{curry-howard-faded}\end{minipage}}
\newcommand{\connection}[1]{$\cdots\includegraphics[height=1em]{#1}\cdots$}
\begin{frame}{Curry--Howard's Rosetta stone}
  \bigskip
  \centering
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{r@{}c@{}l}
    \hil{mathematics} & & \hil{programming} \\
    proving a claim & \connection{emoji-thinking} & implementing a function \\
    stating a claim & \connection{emoji-megaphone} & writing down a type signature \\
    using a lemma & \connection{emoji-link} & calling a function \\
    {\ \ \ \ \,}coming up with an auxiliary lemma & \connection{emoji-lightbulb} & coming up with an auxiliary function \\
    committing a logical fallacy & \connection{emoji-cross} & committing a type error \\
    \pause
    --- & \connection{emoji-explosion} & encountering a runtime bug
  \end{tabular}
  \pause

@@

\begin{frame}{Three case studies}
  \bigskip
  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \scalebox{0.23}{\sieve{7}{7}}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \justifying
        \textbf{Theorem.}
        For every natural number~$n$, there is a prime larger than~$n$.
      \end{simpleblock}
      \\[0.3em]

@@

  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \includegraphics[width=\textwidth,valign=t]{monad}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \textbf{Theorem.}
        Every infinite sequence~$f : \NN \to \NN$ is \emph{good}
        in that there are numbers $i < j$ such that~$f(i) \leq f(j)$.
      \end{simpleblock}
      \\[0.3em]

      \emph{Proof.} There is a \bad{minimal value}~$f(i)$. Set~$j \defeq i+1$. \qed
    \end{column}
  \end{columns}
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \includegraphics[width=\textwidth,valign=t]{calabi-yau}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \justifying
        \textbf{Theorem.}
        Let~$M$ be a surjective matrix with more rows than columns over a
        commutative ring~$A$. Then~$1 = 0$ in~$A$.
      \end{simpleblock}
      \\[0.3em]

      \justifying\emph{Proof.}
      \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
      The matrix is still surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
      is a contradiction to basic linear algebra. \qed
    \end{column}
  \end{columns}
\end{frame}

@@

- Agda icons
- convenient fictions

- formulation problem
- infinite objects
- proving = programming

>>>>

  \textbf{Theorem.} There is a \hil{unique} best~$L^1$-approximation of~$f$ in~$P_n$.
  \pause

@@

\begin{frame}{Backed by logical metatheorems?}
  \hil{Metatheorems} for backing proof mining have been and are being developed which \ldots
  \begin{enumerate}
    \small
    \item[{\includegraphics[height=1em,valign=c]{emoji-abacus}}] guarantee the
    \hil{extractability} of suitable numerical information in principle \\
    (bounds, convergence rates, moduli of uniqueness, rates of asymptotic
    regularity, \ldots),
    \item[{\includegraphics[height=1em,valign=c]{emoji-robot}}] describe an \hil{algorithm} for carrying out the extraction and
    \item[{\includegraphics[height=1em,valign=c]{emoji-package}}] support \hil{modular treatments} of auxiliary lemmas,
  \end{enumerate}
  \bad{\textbf{provided}} the input proof is \hil{formally} supplied in a
  \hil{certain system}.
  \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/novikov.pdf}{[Kohlenbach--Oliva 2002]}
  \pause

  \bad{\textbf{In practice (2025):}} Algorithms \bad{\textbf{not}} used, only followed
  as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
  \hil{Let us explore tool support!}
  \pause

  NB: The \hil{quality} of the extracted data depends on the
  \hil{sophistication} of the \hil{logical~principles} used in the proof
  (Heine--Borel, Bolzano--Weierstraß, \ldots).
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-0em}\includegraphics[width=\paperwidth]{curry-howard-faded}\end{minipage}}
\newcommand{\connection}[1]{$\cdots\includegraphics[height=1em]{#1}\cdots$}
\begin{frame}{Curry--Howard's Rosetta stone}
  \bigskip
  \centering
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{r@{}c@{}l}
    \hil{mathematics} & & \hil{programming} \\
    proving a claim & \connection{emoji-thinking} & implementing a function \\
    stating a claim & \connection{emoji-megaphone} & writing down a type signature \\
    using a lemma & \connection{emoji-link} & calling a function \\
    {\ \ \ \ \,}coming up with an auxiliary lemma & \connection{emoji-lightbulb} & coming up with an auxiliary function \\
    committing a logical error & \connection{emoji-cross} & committing a type error \\
    reasoning circularly & \connection{emoji-cyclone} & getting stuck in an infinite loop \\
    \pause
    --- & \connection{emoji-explosion} & encountering a runtime bug
  \end{tabular}
  \pause

@@

\newcommand{\laterbad}[1]{\only<1>{#1}\only<2->{\bad{#1}}}
\begin{frame}{Three case studies}
  \bigskip
  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \scalebox{0.23}{\sieve{7}{7}}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \justifying
        \textbf{Theorem.}
        For every natural number~$n$, there is a prime larger than~$n$.
        \fixedhref{https://unimath.github.io/agda-unimath/elementary-number-theory.infinitude-of-primes.html}{\formalized}
      \end{simpleblock}
      \\[0.3em]

@@

  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \includegraphics[width=\textwidth,valign=t]{monad}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \textbf{Theorem.}
        Every infinite sequence~$f : \NN \to \NN$ is \emph{good}
        in that there are numbers $i < j$ such that~$f(i) \leq f(j)$.
        \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.ComputationalContent.Dickson.html}{\formalized}
      \end{simpleblock}
      \\[0.3em]

      \emph{Proof.} There is a \laterbad{minimal value}~$f(i)$. Set~$j \defeq i+1$. \qed
    \end{column}
  \end{columns}
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \includegraphics[width=\textwidth,valign=t]{calabi-yau}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \justifying
        \textbf{Theorem.}
        Let~$M$ be a surjective matrix with more rows than columns over a
        commutative ring~$A$. Then~$1 = 0$ in~$A$.
        \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Krull.Dynamical.html\#example}{\formalized}
      \end{simpleblock}
      \\[0.3em]

      \justifying\emph{Proof.}
      \laterbad{Assume not.} Then there is a \laterbad{maximal ideal}~$\mmm$.
      The matrix is still surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
      is a contradiction to basic linear algebra. \qed
    \end{column}
  \end{columns}
\end{frame}

\renewcommand{\insertframeextra}{a}
\begin{frame}{A case study in commutative algebra}
  \begin{simpleblock}
    \textbf{Theorem.}
    Let~$M$ be a surjective matrix with more rows than columns over a
    commutative ring~$A$. Then~$1 = 0$ in~$A$.
    \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Krull.Dynamical.html\#example}{\formalized}
  \end{simpleblock}
  \\[0.3em]

  \justifying\emph{Proof.}
  \bad{Assume not.} Then there is a \bad{maximal ideal}~$\mmm$.
  The matrix is still surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
  is a contradiction to basic linear algebra. \qed

  {\centering\scalebox{0.9}{\begin{tikzpicture}
    \node (0) at (0,1) {$(0) = \{0\}$};
    \node (1) at (0,5) {$(1) = \ZZ$};
    \node (2) at (-2,4) {$(2)$};
    \node [right of=2] (3) {$(3)$};
    \node [below of=2] (4) {$(4)$};
    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
    \node [right of=3] (5) {$(5)$};
    \node [right of=5] (7) {$(7)$};
    \node [right of=7] (7d) {$\ldots$\phantom{(}};
    \node [right of=7d, xshift=1cm, yshift=-1cm] (max)
    {\small\it maximal among the proper ideals};
    \node [below of=4] (8) {$(8)$};
    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
    \draw (0) -- (8);
    \draw (0) -- (8d);
    \draw (0) -- (6);
    \draw (2) -- (1);
    \draw (3) -- (1);
    \draw (5) -- (1);
    \draw (7) -- (1);
    \draw (7d) -- (1);
    \draw (4) -- (2);
    \draw (8) -- (4);
    \draw (6) -- (2);
    \draw (6) -- (3);
    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple!30]
        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}}\par}
  \pause

  \begin{mining}[0.8\textwidth]\emph{Proof.}
  Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
  have~$u, v \in A$ with
  $
    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
  $ and $
    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right)
  $.
  Hence
  $
    1 = (vy)(ux) = (uy)(vx) = 0
  $. \qed
  \end{mining}
\end{frame}

\addtocounter{framenumber}{-1}
\renewcommand{\insertframeextra}{b}
\begin{frame}{Maximal ideals as convenient fictions?}
  \begin{itemize}
    \item[\triang] In \bad{classical mathematics}, every ring has a maximal ideal
    by \bad{Zorn's lemma}.
    \bigskip\pause

    \item[\triang] Without \bad{Zorn}, at least every \bad{countable} ring
    $A = \{ x_0, x_1, \ldots \}$ has a maximal ideal.

    \small
    -- Iterative construction given
    an ideal membership test
    \fixedhref{https://link.springer.com/article/10.1007/BF01454872}{[Krull
    1929]}:
    \begin{align*}
      \mmm_0 &= \{ 0 \}, &
      \mmm_{n+1} &= \begin{cases}
        \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, \\
        \mmm_n, & \text{else.}
      \end{cases}
      \intertext{\visible<3->{-- Also without membership test!
      \fixedhref{https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=a7ff57cacdad6735040459dcf450b231948dbb2f}{[Krivine
      1996]},
      \fixedhref{https://www.sciencedirect.com/science/article/pii/S0168007204000181}{[Berardi--Valentini
      2004]}}}
      \action<3->{\mmm_0 &= \{ 0 \}, &
      \mmm_{n+1} &= \mmm_n + (\underbrace{\{ x_n \,|\, 1 \not\in \mmm_n + (x_n)
      \}}_{\text{a certain subsingleton set}})}
    \end{align*}
    \bigskip
    \pause
    \pause

    \normalsize
    \justifying
    \item[\triang] Without \bad{Zorn}, \hil{every} ring has a maximal ideal in
    a ``suitable \hil{forcing extension} of the universe''
    \fixedhref{https://raw.githubusercontent.com/iblech/constructive-maximal-ideals/master/tex/extended.pdf}{[B.--Schuster
    2024]}.
    In plain terms:
    \hil{Approximate} a (perhaps non-existing) surjection~$\NN \twoheadrightarrow A$
    by partial functions which can \hil{grow on demand}.
  \end{itemize}
\end{frame}
\renewcommand{\insertframeextra}{}

@@
